---
title: "Data storage"
---



```{r}
#| echo: false

library(readr)
library(dplyr)
library(ggplot2)
library(kableExtra)
library(tidyr)
library(scales)
library(tibble)
library(gt)

```

# Raster data types


## Raster data types

- Raster data [on disk]{.underline} is always stored as numeric (no strings, no factors)
- Many numeric data types exist — choice affects **file size** and **precision**
- GDAL powers most raster software → same types across R, Python, QGIS, …

| Data type          |        Minimum |       Maximum | Size | Factor |
| :----------------- | -------------: | ------------: | ---: | -----: |
| Byte               |              0 |           255 |  39M |     1x |
| UInt16             |              0 |        65,535 |  78M |     2x |
| Int16 / CInt16     |        -32,768 |        32,767 |  78M |     2x |
| UInt32             |              0 | 4,294,967,295 | 155M |    ~4x |
| Int32 / CInt32     | -2,147,483,648 | 2,147,483,647 | 155M |    ~4x |
| Float32 / CFloat32 |        -3.4E38 |        3.4E38 | 155M |    ~4x |
| Float64 / CFloat64 |      -1.79E308 |      1.79E308 | 309M |    ~8x |

: Numeric data types supported by GDAL [source: @amatulli2024]. File size using a constant dataset (same values and resolution) {#tbl-datatypes}



:::{.notes}

- Categorical data → use integer type and store the label–value mapping in the metadata
- Always use the **smallest** datatype that covers your value range
- Q: values 0–1 (fraction)? → not `Float32`! Multiply by 100 → `Byte`, or ×1000 → `UInt16`
- Q: temperature –20 to +40°C? → not `CFloat32`! Multiply by 100 → `CInt16` (–2000 to 4000)
- Q: NDVI –1 to 1? → rescale to 0–255 → `Byte`

:::


## Choosing a data type

- Naive approach: let the value range *dictate* the datatype (e.g. NDVI –1 to 1 → `Float32`)
- Better approach: **transform** values to fit a smaller datatype

. . .

| Data     | Data Range    | Naive type | Smart type | New Range        |
| :------- | ------------- | :--------- | :--------- | ---------------- |
| Fraction | `0`-`1`       | `Float32`  | `Byte`     | `0` - `100`      |
| NDVI     | `-1`-`+1`     | `CFloat32` | `Byte`     | `0` - `255`      |
| Temp     | `-20` - `+40` | `CFloat32` | `CInt16`   | `–2000` - `4000` |


:::{.notes}

- The precision lost by using a smaller type is often negligible for the application
- GDAL stores scale/offset in file metadata, so any GDAL-powered software reverses the transform on read

:::


## Min-max rescaling

General formula to rescale $x$ into range $[a, b]$:

$$x' = a + \frac{(x - \min(x))\,(b - a)}{\max(x) - \min(x)}$$

For NDVI ($x \in [-1,\,1]$, target $[a, b] = [0,\,255]$):

. . .

$$x' = 0 + \frac{(x - (-1))\,\times\,255}{1-(-1)} = \frac{(x+1)\times 255}{2}$$

. . . 

Simplified:

$$\boxed{x' = 127.5\,x + 127.5}$$

→ e.g. NDVI 0.2 is stored as $127.5 \times 0.2 + 127.5 = 153$

:::{.notes}

- $a = 0$, $b = 255$ (Byte range); $\min(x) = -1$, $\max(x) = 1$ (NDVI range)
- The two parameters 127.5 and 127.5 are the `scale` and `offset` of this linear transform

:::


## Precision trade-off

Storing NDVI in `Byte` limits us to 255 distinct values:

$$\text{precision} = \frac{\max(x) - \min(x)}{b - a} = \frac{1 - (-1)}{255 - 0} \approx 0.0078$$

→ Every stored NDVI is rounded to the nearest 0.0078

. . .

For most vegetation mapping applications, NDVI differences below ~0.01 are not ecologically meaningful → `Byte` precision is usually sufficient

. . .

Need finer precision? Use `UInt16` (0–65,535): $\frac{2}{65535} \approx 0.00003$


## R: rescaling vectors


```{r}
#| echo: true

scale_minmax <- function(x, a = 0, b = 255) {
  a + (x - min(x)) * (b - a) / (max(x) - min(x))
}

measured_ndvi <- seq(-1, 1, 0.1)
stored_value <- scale_minmax(measured_ndvi)
```


```{r}
#| echo: false


tibble(measured_ndvi, stored_value) |>
  ggplot(aes(measured_ndvi, stored_value)) +
  geom_line(col = "lightgrey") +
  geom_point() +
  labs(x = "Measured NDVI (-1 to +1)", y = "Stored value (0 to 255)") + 
  scale_y_continuous(breaks = c(seq(0, 255, 50))) +
  theme_minimal() 
```






## Restoring original values

- The transform $x' = 127.5\,x + 127.5$ is reversible
- Invert the function to recover originals: $x = \dfrac{x' - 127.5}{127.5}$
- More generally: two parameters describe the transform — `scale` and `offset`
- *Ideally* transformation of values is not done within the R / Python session. Best practice:
  - GDAL transforms values during writing
  - GDAL stores scale and offset in the **file metadata** 
  - GDAL restores the original values on read
- However, this is not always the case "in the wild"

:::{.notes}

- No need to tell your colleague about the transform — it's in the GeoTIFF metadata
- `rast("INT1U.tif")` returns the original floating-point values; the Byte storage is invisible

:::


## GDAL scale/offset in R 

- `terra::writeRaster` supports `scale =` and `offset =` arguments — it transforms on write, inverts on read.
- The user needs to know (calculate) these values beforehand
- The formula to determine `scale` and `offset` can be derived by the min-max formula

$$\text{scale} = \frac{b - a}{\max(x) - \min(x)}, \qquad \text{offset} = \frac{a\,\max(x) - b\,\min(x)}{\max(x) - \min(x)}$$

---


```{r}
#| echo: true

get_scale_offset <- function(x, a = 0, b = 255) {  # x must be a SpatRaster
  library(terra)
  mm     <- minmax(x)                # <3>
  min_x  <- mm[1, 1]
  max_x  <- mm[2, 1]

  scale  <- (b - a) / (max_x - min_x)
  offset <- (a * max_x - b * min_x) / (max_x - min_x)

  list(
    scale          = scale,
    offset         = offset,
    scale_inverted  = 1 / scale,      # <1>
    offset_inverted = offset * -1     # <2>
  )
}
```

1. `writeRaster` *divides* by scale (!)
2. `writeRaster` *subtracts* offset (!)
3. Use the global min/max function, not per-cell (e.g. `min()`)


:::{.notes}

- `min(raster)` returns per-cell minimum over bands, NOT the global minimum → always use `minmax()` or `global()`
- The inversion is necessary because writeRaster's convention is: stored_value = x / scale - offset (divide first, then subtract)

:::


## Scale / offset in practice

```{r}
#| echo: true
library(terra)

elev <- rast(system.file("ex/elev.tif", package = "terra"))
plot(elev, main = paste(minmax(elev), collapse = " – "))
```

---

```{r}
#| echo: true

so <- get_scale_offset(elev)

so

# Note: floating-point rounding may introduce NAs at boundary values
writeRaster(elev, "data-out/INT1U.tif", datatype = "INT1U", overwrite = TRUE,
            scale = so$scale_inverted, offset = so$offset_inverted)

writeRaster(elev, "data-out/FLT4S.tif", datatype = "FLT4S", overwrite = TRUE)


```

```{r}
file.size("data-out/INT1U.tif") / file.size("data-out/FLT4S.tif")
```

---

Scale and offset are visible in the file metadata — any GDAL-powered tool can read them:

```{sh}
gdalinfo data-out/INT1U.tif
```

:::{.notes}

- Reading `INT1U.tif` back gives original float values — the Byte storage is transparent
- The ratio shows how much smaller the Byte file is compared to Float32 (`FLT4S`)
- In the gdalinfo output, look for `Type=Byte` (stored type) and `Scale`/`Offset` under Band 1

:::


# NoData in rasters

## NoData in rasters

- Raster files have **no explicit null type** — every cell must hold a numeric value
- To mark cells as missing, a two-step approach is used:
  1. Assign a **reserved value** within the datatype's range (e.g. the maximum possible value)
  2. **Label** that value as `NoData` in the file metadata

. . .

→ Any software reading the file interprets that reserved value as missing, not as a real measurement

:::{.notes}

- This is different from R's in-memory `NA`, which is a proper null type
- The reserved value label lives in the metadata, just like scale/offset

:::


## NoData & data types

For **integer types**, the reserved value must fit within the datatype's range:

| Datatype        | Common NoData reserved value |
| :-------------- | :--------------------------- |
| `Byte`          | 255 (max value)              |
| `Int16`         | −32,768 (min value)          |
| `Float32`[^NAN] | NaN or −9999                 |


[^NAN]: Float types support NaN (as part of the IEEE 754 standard). `-9999` sometimes seen in float rasters is a legacy workaround for older software that cannot handle NaN



. . .

**Risk (integer types):** if real data happens to contain the reserved value → those cells silently become `NoData`

:::{.notes}

- For Byte: using 255 as NoData means you only have 0–254 for real data
- NaN is an unambiguous bit pattern in IEEE 754 — it represents "undefined result" (e.g. 0/0) and is never a valid number
- The silent data-loss risk applies only to integer types; float types with NaN are safe

:::


## NoData in terra

Inspect and set the NoData flag:

```{r}
#| echo: true
library(terra)

elev <- rast(system.file("ex/elev.tif", package = "terra"))

NAflag(elev)   # current NoData reserved value
```

---

```{r}
#| echo: true
#| fig-cap: "Note that the maximum elevation values are now NA"
#| classes: .center
# write with an explicit NoData flag
writeRaster(elev, "data-out/elev_naflag.tif", datatype = "INT1U", overwrite = TRUE,
            scale = so$scale_inverted, offset = so$offset_inverted)            

# read back — terra interprets 255 as NA automatically
elev_back <- rast("data-out/elev_naflag.tif")
plot(elev_back, colNA = "black")
```

---

The reserved value and its NoData label are both visible in the metadata:

```{sh}
gdalinfo data-out/elev_naflag.tif
```

:::{.notes}

- `NAflag()` returns the value currently flagged as NoData; `NA` means no flag is set
- On read, terra converts the flagged value back to R's `NA` transparently
- `NAflag` and `datatype` are set together in `writeRaster` — they are two sides of the same coin
- In the gdalinfo output, look for `NoData Value=255` under Band 1

:::


# Categorical rasters


## Categorical rasters

- Some raster data is **categorical**, not continuous: land use, soil type, vegetation class, …
- Raster files cannot store text — categories are encoded as **integers**
- The mapping (integer code → category label) is stored in the file metadata as an **attribute table**

. . .

→ Analogous to R's `factor`: an underlying integer + a levels table

. . .

→ Choose the smallest integer type that covers the number of categories — `Byte` handles up to 255 classes

:::{.notes}

- Most land use / land cover classifications have far fewer than 255 classes → Byte is almost always sufficient
- The attribute table lives in the file metadata, just like scale/offset and NoData flags

:::


## Categorical rasters in terra

```{r}
#| echo: true
elev_cat <- classify(elev, c(0, 200, 350, 550))                                     # <1>
levels(elev_cat) <- data.frame(value = 0:2, label = c("low", "medium", "high"))     # <2>

plot(elev_cat)
```

1. Reclassify values into three categories (`0-200`, `200-350` and `350-550`)
2. Relable the default classes to `low`, `medium` and `high`

---

```{r}
#| echo: true
# write as Byte explicitly (optional, writeRaster makes a reasonable guess)
writeRaster(elev_cat, "data-out/elev_cat.tif", datatype = "INT1U", overwrite = TRUE)
```

[]{.gap}

```{sh}
gdalinfo data-out/elev_cat.tif
```
:::{.notes}

- `levels()` sets the attribute table — same function as for R factors
- `cats(elev_cat)` to inspect the table after the fact
- On read, terra restores the levels automatically if the attribute table is embedded in the file

:::


```{r}
#| echo: false

filesize2df <- function(file, comment){
  data.frame(
    comment  = comment,
    size_mb  = round(file.size(file) / 1024^2, 2)
  )
}

bind_print <- function(...){
  rbind(...) |>
    dplyr::mutate(difference = paste(round((size_mb / size_mb[1] - 1) * 100), "%")) |>
    knitr::kable(col.names = c("File", "Size (MB)", "Difference"), align = "lrr")
}
```

# Compression

## Raster file size

- File size = **dimensions × bands × bytes per pixel**
- Example: single SRTM tile, one band, `Int16` (2 bytes/pixel):

$$3601 \times 3601 \times 2\,\text{B} = 25{,}934{,}402\,\text{B} \approx 25.9\,\text{MB}$$

- **Compression** reduces storage by exploiting spatial redundancy in pixel values

:::{.notes}

- Real-world rasters often have large homogeneous regions → high redundancy → compress well
- Redundancy is especially high in categorical rasters (e.g. land cover with few distinct values)

:::


## Lossless vs lossy

|                 | **Lossless**                         | **Lossy**                     |
| :-------------- | :----------------------------------- | :---------------------------- |
| Data preserved? | Perfectly                            | Approximated                  |
| Use case        | Scientific data (DEM, multispectral) | Photographic / visual imagery |
| Examples        | LZW, DEFLATE, ZSTD, PACKBITS         | JPEG2000, WEBP                |

. . .

→ For scientific analysis **always use lossless** — lossy compression permanently alters pixel values

:::{.notes}

- WEBP supports both lossy and lossless modes depending on settings
- Once lossy-compressed, original values cannot be recovered — important for analysis reproducibility
- Much of this section is based on @gandhi2020 and [Paul Ramsey's GeoTIFF compression guide](https://blog.cleverelephant.ca/2015/02/geotiff-compression-for-dummies.html)

:::


## How compression works

Exploit **repeated values** — store each unique value once, track its positions:

|            |                                                                                         |
| :--------- | :-------------------------------------------------------------------------------------- |
| Original   | `100, 101, 102, 100, 100`                                                               |
| Compressed | `100` → positions `[0, 3, 4]` <br> `101` → positions `[1]` <br> `102` → positions `[2]` |

→ 5 values encoded as 3 unique values + a position list

. . .

Works best when data contains **many repeated or near-repeated values** (flat terrain, uniform land cover)

:::{.notes}

- This is a simplified illustration of dictionary / run-length encoding
- LZW and DEFLATE use more sophisticated variants of this principle

:::


## PREDICTOR

LZW, DEFLATE, and ZSTD can use a **predictor** to improve compression further:

→ Store **differences** between consecutive values instead of absolute values

|                |        |       |       |       |       |
| :------------- | :----- | ----- | ----- | ----- | ----- |
| Original       | `100`  | `101` | `102` | `100` | `100` |
| With predictor | `100`  | `+1`  | `+1`  | `−2`  | `0`   |

Small differences compress better than large absolute values

. . .

| Option        | Use for                  |
| :------------ | :----------------------- |
| `PREDICTOR=1` | No predictor (default)   |
| `PREDICTOR=2` | Integer data (e.g. DEMs) |
| `PREDICTOR=3` | Float data               |

:::{.notes}

- Predictor=2 works well for integer rasters with gradual spatial variation
- Predictor=3 for float rasters
- Using the wrong predictor (e.g. =3 on integer data) still works but may not compress as well

:::


## Tiling & compression costs

- By default, GeoTIFF stores data **line by line** (strip layout)
- `TILED=YES` stores data in **256 × 256 pixel blocks**

→ Tiling improves random-access reads and is essential for **Cloud-Optimized GeoTIFF (COG)**

. . .

**Compression is not free:**

- Adds CPU overhead on **write and read**
- Good default for most scientific workflows: **`DEFLATE` + `TILED=YES`**

:::{.notes}

- Strip layout is fast for sequential top-to-bottom reads but slow for spatial queries on large files
- ZSTD generally offers better ratios than DEFLATE but has less universal software support
- COG = Cloud-Optimized GeoTIFF; tiling + overviews enable efficient streaming from cloud storage

:::


## gdal_translate

[`gdal_translate`](https://gdal.org/en/stable/programs/gdal_translate.html) converts raster files between formats and applies compression:

```{bash}
gdal_translate --help
```

Key points:

- Two **positional arguments**: `<input_file>` `<output_file>`
- Compression via `-co KEY=VALUE` (**c**reation **o**ptions)
- Options are format-specific — always check `--help`

:::{.notes}

- Snake case (`gdal_translate`) vs no separator (`gdalinfo`) — GDAL naming is inconsistent, always verify
- Mutually exclusive options are shown as `[[-strict]|[-not_strict]]` in the help output

:::


## gdal_translate: baseline

```{bash}
#| eval: false
gdal_translate data/raster-deepdive/dhm25_grid_raster.asc data-out/dhm25.tif
```

```{r}
#| echo: false
size0 <- filesize2df("data/raster-deepdive/dhm25_grid_raster.asc", "Original (ASCII)")
size1 <- filesize2df("data-out/dhm25.tif", "GeoTIFF, no compression")
bind_print(size0, size1)
```


## gdal_translate: + DEFLATE

```{bash}
#| eval: false
gdal_translate data/raster-deepdive/dhm25_grid_raster.asc data-out/dhm25_2.tif \
  -co COMPRESS=DEFLATE
```

```{r}
#| echo: false
size2 <- filesize2df("data-out/dhm25_2.tif", "`COMPRESS=DEFLATE`")
bind_print(size0, size1, size2)
```


## gdal_translate: + TILED

```{bash}
#| eval: false
gdal_translate data/raster-deepdive/dhm25_grid_raster.asc data-out/dhm25_3.tif \
  -co COMPRESS=DEFLATE \
  -co TILED=YES
```

```{r}
#| echo: false
size3 <- filesize2df("data-out/dhm25_3.tif", "`TILED=YES`")
bind_print(size0, size1, size2, size3)
```


## gdal_translate: + PREDICTOR

```{bash}
#| eval: false
gdal_translate data/raster-deepdive/dhm25_grid_raster.asc data-out/dhm25_4.tif \
  -co COMPRESS=DEFLATE \
  -co TILED=YES \
  -co PREDICTOR=3
```

```{r}
#| echo: false
size4 <- filesize2df("data-out/dhm25_4.tif", "`PREDICTOR=3`")
bind_print(size0, size1, size2, size3, size4)
```

:::{.notes}

- `PREDICTOR=3` for floating-point data; use `PREDICTOR=2` for integer data
- Each row in the table adds one `-co` flag — the cumulative effect shows how the options interact

:::


# Bringing it all together

## Three tools, one goal

All three strategies reduce on-disk file size — but they work at **different levels**:

| Tool | Targets | Mechanism |
|:-----|:--------|:----------|
| **Datatype** | Per-pixel cost | 1 byte (`Byte`) vs 4 bytes (`Float32`) |
| **Scale / offset** | Value range | Rescale floats to fit a smaller integer type |
| **Compression** | Spatial redundancy | Exploit repeated values across pixels |

. . .

→ They are **complementary** — apply all three for maximum effect

→ Order matters: choose datatype first, then compress

:::{.notes}

- Getting the datatype wrong means you compress a larger file to begin with
- Scale/offset is what makes integer types viable for continuous data — without it you'd lose the value range
- Compression then removes the spatial redundancy that remains regardless of datatype

:::


## Combined example

Rows 1 & 2 reuse the `dhm25` variants from the compression section; rows 3 & 4 add scale/offset:

```{r}
#| echo: true
library(terra)

dhm <- rast("data/raster-deepdive/dhm25_grid_raster.asc")
so  <- get_scale_offset(dhm)

# reuses dhm25.tif (baseline) and dhm25_4.tif (best compression) from earlier

# 3. Byte + scale/offset, no compression
writeRaster(dhm, "data-out/dhm25_byte.tif",
            datatype = "INT1U", overwrite = TRUE,
            scale = so$scale_inverted, offset = so$offset_inverted)

# 4. Byte + scale/offset + compression
writeRaster(dhm, "data-out/dhm25_byte_comp.tif",
            datatype = "INT1U", overwrite = TRUE,
            scale = so$scale_inverted, offset = so$offset_inverted,
            gdal = c("COMPRESS=DEFLATE", "TILED=YES"))
```

---

```{r}
#| echo: false
s0 <- filesize2df("data-out/dhm25.tif",           "GeoTIFF, no compression (baseline)")
s1 <- filesize2df("data-out/dhm25_4.tif",         "DEFLATE + TILED + PREDICTOR")
s2 <- filesize2df("data-out/dhm25_byte.tif",      "Byte + scale/offset")
s3 <- filesize2df("data-out/dhm25_byte_comp.tif", "Byte + scale/offset + DEFLATE + TILED")
bind_print(s0, s1, s2, s3)
```

:::{.notes}

- Row 2 shows what compression alone buys
- Row 3 shows what datatype reduction alone buys (scale/offset makes it possible)
- Row 4 shows the multiplicative effect of combining both strategies
- The two strategies are not redundant — each contributes independently

:::


## Practical checklist

When writing a raster to disk:

1. **What is my value range?** → choose the smallest datatype that fits
2. **Can I rescale to fit an integer type?** → use `scale` / `offset` in `writeRaster`
3. **Are there missing cells?** → set `NAflag` (integer types) or use NaN (float types)
4. **Apply lossless compression:** `COMPRESS=DEFLATE` + `TILED=YES`
   - Add `PREDICTOR=2` for integer data, `PREDICTOR=3` for float data

. . .

**Default recommendation for most scientific rasters:**

```
INT1U / Int16  +  scale/offset  +  DEFLATE  +  TILED=YES  +  PREDICTOR=2
```

:::{.notes}

- For data that genuinely needs float precision (e.g. model outputs with many significant digits), skip the scale/offset step and use FLT4S + DEFLATE + TILED + PREDICTOR=3
- ZSTD can replace DEFLATE for better ratios where broad software compatibility is less critical

:::


## References