---
title: "Compression"
engine: knitr
---

```{r}
#| echo: false

filesize2df <- function(file, comment){
  data.frame(
    comment  = comment,
    size_mb  = round(file.size(file) / 1024^2, 2)
  )
}

bind_print <- function(...){
  rbind(...) |>
    dplyr::mutate(difference = paste(round((size_mb / size_mb[1] - 1) * 100), "%")) |>
    knitr::kable(col.names = c("File", "Size (MB)", "Difference"), align = "lrr")
}
```

# Compression

## Raster file size

- File size = **dimensions × bands × bytes per pixel**
- Example: single SRTM tile, one band, `Int16` (2 bytes/pixel):

$$3601 \times 3601 \times 2\,\text{B} = 25{,}934{,}402\,\text{B} \approx 25.9\,\text{MB}$$

- **Compression** reduces storage by exploiting spatial redundancy in pixel values

:::{.notes}

- Real-world rasters often have large homogeneous regions → high redundancy → compress well
- Redundancy is especially high in categorical rasters (e.g. land cover with few distinct values)

:::


## Lossless vs lossy

| | **Lossless** | **Lossy** |
|:--|:------------|:---------|
| Data preserved? | Perfectly | Approximated |
| Use case | Scientific data (DEM, multispectral) | Photographic / visual imagery |
| Examples | LZW, DEFLATE, ZSTD, PACKBITS | JPEG2000, WEBP |

. . .

→ For scientific analysis **always use lossless** — lossy compression permanently alters pixel values

:::{.notes}

- WEBP supports both lossy and lossless modes depending on settings
- Once lossy-compressed, original values cannot be recovered — important for analysis reproducibility
- Much of this section is based on @gandhi2020 and [Paul Ramsey's GeoTIFF compression guide](https://blog.cleverelephant.ca/2015/02/geotiff-compression-for-dummies.html)

:::


## How compression works

Exploit **repeated values** — store each unique value once, track its positions:

| | |
|:--|:--|
| Original | `100, 101, 102, 100, 100` |
| Compressed | `100` → positions `[0, 3, 4]`; `101` → `[1]`; `102` → `[2]` |

→ 5 values encoded as 3 unique values + a position list

. . .

Works best when data contains **many repeated or near-repeated values** (flat terrain, uniform land cover)

:::{.notes}

- This is a simplified illustration of dictionary / run-length encoding
- LZW and DEFLATE use more sophisticated variants of this principle

:::


## PREDICTOR

LZW, DEFLATE, and ZSTD can use a **predictor** to improve compression further:

→ Store **differences** between consecutive values instead of absolute values

| | Values |
|:--|:-------|
| Original | `100, 101, 102, 100, 100` |
| With predictor | `100, +1, +1, −2, 0` |

Small differences compress better than large absolute values

. . .

| Option | Use for |
|:-------|:--------|
| `PREDICTOR=1` | No predictor (default) |
| `PREDICTOR=2` | Integer data (e.g. DEMs) |
| `PREDICTOR=3` | Float data |

:::{.notes}

- Predictor=2 works well for integer rasters with gradual spatial variation
- Predictor=3 for float rasters
- Using the wrong predictor (e.g. =3 on integer data) still works but may not compress as well

:::


## Tiling & compression costs

- By default, GeoTIFF stores data **line by line** (strip layout)
- `TILED=YES` stores data in **256 × 256 pixel blocks**

→ Tiling improves random-access reads and is essential for **Cloud-Optimized GeoTIFF (COG)**

. . .

**Compression is not free:**

- Adds CPU overhead on **write and read**
- Good default for most scientific workflows: **`DEFLATE` + `TILED=YES`**

:::{.notes}

- Strip layout is fast for sequential top-to-bottom reads but slow for spatial queries on large files
- ZSTD generally offers better ratios than DEFLATE but has less universal software support
- COG = Cloud-Optimized GeoTIFF; tiling + overviews enable efficient streaming from cloud storage

:::


## gdal_translate

[`gdal_translate`](https://gdal.org/en/stable/programs/gdal_translate.html) converts raster files between formats and applies compression:

```{bash}
gdal_translate --help
```

Key points:

- Two **positional arguments**: `<input_file>` `<output_file>`
- Compression via `-co KEY=VALUE` (**c**reation **o**ptions)
- Options are format-specific — always check `--help`

:::{.notes}

- Snake case (`gdal_translate`) vs no separator (`gdalinfo`) — GDAL naming is inconsistent, always verify
- Mutually exclusive options are shown as `[[-strict]|[-not_strict]]` in the help output

:::


## gdal_translate: baseline

```{bash}
gdal_translate data/week5-exercises/dhm25_grid_raster.asc data-out/dhm25.tif
```

```{r}
#| echo: false
size0 <- filesize2df("data/week5-exercises/dhm25_grid_raster.asc", "Original (ASCII)")
size1 <- filesize2df("data-out/dhm25.tif",                         "GeoTIFF, no compression")
bind_print(size0, size1)
```


## gdal_translate: + DEFLATE

```{bash}
gdal_translate data/week5-exercises/dhm25_grid_raster.asc data-out/dhm25_2.tif \
  -co COMPRESS=DEFLATE
```

```{r}
#| echo: false
size2 <- filesize2df("data-out/dhm25_2.tif", "`COMPRESS=DEFLATE`")
bind_print(size0, size1, size2)
```


## gdal_translate: + TILED

```{bash}
gdal_translate data/week5-exercises/dhm25_grid_raster.asc data-out/dhm25_3.tif \
  -co COMPRESS=DEFLATE \
  -co TILED=YES
```

```{r}
#| echo: false
size3 <- filesize2df("data-out/dhm25_3.tif", "`TILED=YES`")
bind_print(size0, size1, size2, size3)
```


## gdal_translate: + PREDICTOR

```{bash}
gdal_translate data/week5-exercises/dhm25_grid_raster.asc data-out/dhm25_4.tif \
  -co COMPRESS=DEFLATE \
  -co TILED=YES \
  -co PREDICTOR=3
```

```{r}
#| echo: false
size4 <- filesize2df("data-out/dhm25_4.tif", "`PREDICTOR=3`")
bind_print(size0, size1, size2, size3, size4)
```

:::{.notes}

- `PREDICTOR=3` for floating-point data; use `PREDICTOR=2` for integer data
- Each row in the table adds one `-co` flag — the cumulative effect shows how the options interact

:::
