---
title: "Raster data on disk"
subtitle: "Spatiotemporal Data Science · FS26"
---

```{r}
#| echo: false

library(readr)
library(dplyr)
library(ggplot2)
library(kableExtra)
library(tidyr)
library(scales)
library(tibble)
library(gt)
```


## Today

:::{.incremental}
1. **GDAL CLI** — what it is, when to use it, how commands are structured
1. **Raster data types** — choosing the right type, rescaling, scale/offset
1. **gdal raster info** — inspecting raster files from the terminal
1. **NoData** — how missing values are stored on disk
1. **Categorical rasters** — integer encoding + attribute tables
1. **Compression** — lossless compression with `gdal raster convert`
1. **Reprojection** — changing CRS with `gdal raster reproject`
:::


# GDAL CLI {background-color="#1269B0"}


## GDAL CLI

- GDAL is the basis for most GIS software — we can also call it **directly from the terminal**
- Ships several [raster programs](https://gdal.org/en/stable/programs/#raster-programs), each designed for a specific task
- Advantages over R / Python: faster, memory-efficient, composable, portable
- Install via conda:

```{.bash}
conda install -c conda-forge gdal
``` 

:::{.notes}

- We've already used GDAL indirectly: terra, QGIS, and many other tools call it under the hood
- "Block-by-block" means GDAL reads and writes in tiles — a 10GB raster can be processed with <1MB RAM
- In practice: use CLI for data prep / inspection, R/Python for analysis and visualization

:::


## When does GDAL CLI shine?

:::{.incremental}
- **Reprojection**: Block-by-block — no full load into RAM → fast!
- **Format conversion**: no full load into RAM → fast!
- **Extracting file info**: Instant, no environment startup — great for quick access
- **Batch processing**: One shell loop over hundreds of files; no memory accumulation
- **Mosaicking**: Combining hundreds of tiles stays out of RAM entirely
- **Building overviews**: Purely structural file operation
- **Pipelines / CI**: Slots into shell scripts, Makefiles, CI/CD without an R/Python runtime
:::

:::{.notes}

- Reprojection and batch processing are the two most commonly cited by practitioners
- "No environment startup" matters a lot in scripts: starting R just to run gdalinfo on 1000 files adds minutes of overhead
- Overviews (pyramids) allow fast display at lower resolutions — QGIS and other tools use them automatically
- We'll have a look at *reprojection* and *file info extraction*
:::

## Which CLI tools do you already use?

. . .

- `conda` — manage environments and packages
- `pip` — install Python packages
- `git` — version control


:::{.notes}

- Ask students to raise their hand for each one — most will have used at least 3
- The point: they're already CLI users; GDAL is not a new paradigm, just a new tool
- Some may also use `ssh`, `jupyter`, `docker` — worth asking

:::


## Anatomy of a CLI call

```
program    subcommand    flag  flag-value     argument
   ↓           ↓          ↓       ↓             ↓
conda     install         -c   conda-forge     gdal
⎺⎺⎺     ⎺⎺⎺⎺⎺       ⎺⎺  ⎺⎺⎺⎺⎺⎺⎺     ⎺⎺
```

. . .

- **Program** — the tool being called
- **Subcommand** — the action to perform (not all tools have these)
- **Flag** — modifies behaviour; `-c` is shorthand for `--channel`
- **Flag value** — the value passed to a flag
- **Argument** — positional value the command operates on (no `-` prefix)

:::{.notes}

- Short flags (`-c`) and long flags (`--channel`) are equivalent; long flags are more readable in scripts
- Flags can appear in any order; arguments are positional (order matters)
- This anatomy applies to every CLI tool: `git commit -m "message"`, `quarto render --to pdf`, `gdal raster info -f json file.tif`

:::


## New GDAL API

- Prior to GDAL 3.11, the suite of GDAL programs had a highly inconsistent API
- With 3.11, GDAL introduced a unified CLI entry point — `gdal <noun> <verb>`
- The new API is great, but WIP
- Our examples use the new API
- See webinar [GDAL Command Line Interface Modernization ](https://youtu.be/ZKdrYm3TiBU?si=IAmioD_tCOhKgLyu)


| Task | Old API | New API (3.11+) |
|:-----|:--------|:----------------|
| Inspect | `gdalinfo` | `gdal raster info` |
| Convert | `gdal_translate` | `gdal raster convert` |
| Reproject | `gdalwarp` | `gdal raster reproject` |


:::{.notes}

- You will see the old API in the wild — documentation, Stack Overflow answers often still use it
- The old tools are not deprecated; new API is additive
- Check your version: `gdal --version`

:::


## Input → output

GDAL is **non-destructive by design** — it always reads from a source and writes to a new destination:

```
gdal raster reproject --dst-crs EPSG:2056 input.tif output.tif
                                           ─────────  ──────────
                                           source     new file, never touches source
```

. . .

- The original file is **never modified**
- Every operation requires both `<input>` and `<output>` arguments
- Consequence: you need roughly **2× the disk space** during processing

. . .

**Exception:** `gdal edit` — metadata-only changes (CRS, NoData, scale/offset) can be applied in place

:::{.notes}

- Non-destructive is safer: if something goes wrong, the original is intact
- `gdal edit` is useful when you just need to assign a missing CRS or fix a NoData flag without rewriting the full file
- In pipelines, the output of one step becomes the input of the next — plan your disk usage accordingly

:::


# Raster data types {background-color="#1269B0"}


## Raster data types

- Raster data [on disk]{.underline} is always stored as numeric (no strings, no factors)
- Many numeric data types exist — choice affects **file size** and **precision**
- GDAL powers most raster software → same types across R, Python, QGIS, …

| Data type          |        Minimum |       Maximum | Size | Factor |
| :----------------- | -------------: | ------------: | ---: | -----: |
| Byte               |              0 |           255 |  39M |     1x |
| UInt16             |              0 |        65,535 |  78M |     2x |
| Int16 / CInt16     |        -32,768 |        32,767 |  78M |     2x |
| UInt32             |              0 | 4,294,967,295 | 155M |    ~4x |
| Int32 / CInt32     | -2,147,483,648 | 2,147,483,647 | 155M |    ~4x |
| Float32 / CFloat32 |        -3.4E38 |        3.4E38 | 155M |    ~4x |
| Float64 / CFloat64 |      -1.79E308 |      1.79E308 | 309M |    ~8x |

: Numeric data types supported by GDAL [source: @amatulli2024]. File size using a constant dataset (same values and resolution) {#tbl-datatypes}


:::{.notes}

- Categorical data → use integer type and store the label–value mapping in the metadata
- Always use the **smallest** datatype that covers your value range
- Q: values 0–1 (fraction)? → not `Float32`! Multiply by 100 → `Byte`, or ×1000 → `UInt16`
- Q: temperature –20 to +40°C? → not `CFloat32`! Multiply by 100 → `CInt16` (–2000 to 4000)
- Q: NDVI –1 to 1? → rescale to 0–255 → `Byte`

:::


## Choosing a data type

- Naive approach: let the value range *dictate* the datatype (e.g. NDVI –1 to 1 → `Float32`)
- Better approach: **transform** values to fit a smaller datatype

. . .

| Data     | Data Range    | Naive type | Smart type | New Range        |
| :------- | ------------- | :--------- | :--------- | ---------------- |
| Fraction | `0`-`1`       | `Float32`  | `Byte`     | `0` - `100`      |
| NDVI     | `-1`-`+1`     | `CFloat32` | `Byte`     | `0` - `255`      |
| Temp     | `-20` - `+40` | `CFloat32` | `CInt16`   | `–2000` - `4000` |


:::{.notes}

- The precision lost by using a smaller type is often negligible for the application
- GDAL stores scale/offset in file metadata, so any GDAL-powered software reverses the transform on read

:::


## Min-max rescaling

General formula to rescale $x$ into range $[a, b]$:

$$x' = a + \frac{(x - \min(x))\,(b - a)}{\max(x) - \min(x)}$$

For NDVI ($x \in [-1,\,1]$, target $[a, b] = [0,\,255]$):

. . .

$$x' = 0 + \frac{(x - (-1))\,\times\,255}{1-(-1)} = \frac{(x+1)\times 255}{2}$$

. . .

Simplified:

$$\boxed{x' = 127.5\,x + 127.5}$$

→ e.g. NDVI 0.2 is stored as $127.5 \times 0.2 + 127.5 = 153$

:::{.notes}

- $a = 0$, $b = 255$ (Byte range); $\min(x) = -1$, $\max(x) = 1$ (NDVI range)
- The two parameters 127.5 and 127.5 are the `scale` and `offset` of this linear transform

:::


## Precision trade-off

Storing NDVI in `Byte` limits us to 256 distinct values:

$$\text{precision} = \frac{\max(x) - \min(x)}{b - a} = \frac{1 - (-1)}{255 - 0} \approx 0.0078$$

→ Every stored NDVI is rounded to the nearest 0.0078

. . .

For most vegetation mapping applications, NDVI differences below ~0.01 are not ecologically meaningful → `Byte` precision is usually sufficient

. . .

Need finer precision? Use `UInt16` (0–65,535): $\frac{2}{65535} \approx 0.00003$


## R: rescaling vectors


```{r}
#| echo: true

scale_minmax <- function(x, a = 0, b = 255) {
  x_new <- a + (x - min(x)) * (b - a) / (max(x) - min(x))
  round(x_new, 0)
}

measured_ndvi <- seq(-1, 1, 0.1)
stored_value <- scale_minmax(measured_ndvi)
```


```{r}
#| echo: false


tibble(measured_ndvi, stored_value) |>
  ggplot(aes(measured_ndvi, stored_value)) +
  geom_line(col = "lightgrey") +
  geom_point() +
  labs(x = "Measured NDVI (-1 to +1)", y = "Stored value (0 to 255)") +
  scale_y_continuous(breaks = c(seq(0, 255, 50))) +
  theme_minimal()
```


## Restoring original values

- The transform $x' = 127.5\,x + 127.5$ is reversible
- Invert the function to recover originals: $x = \dfrac{x' - 127.5}{127.5}$
- More generally: two parameters describe the transform — `scale` and `offset`
- *Ideally* transformation of values is not done within the R / Python session. Best practice:
  - GDAL transforms values during writing
  - GDAL stores scale and offset in the **file metadata**
  - GDAL restores the original values on read
- However, this is not always the case "in the wild"

:::{.notes}

- No need to tell your colleague about the transform — it's in the GeoTIFF metadata
- `rast("INT1U.tif")` returns the original floating-point values; the Byte storage is invisible

:::


## GDAL scale/offset in R

- `terra::writeRaster` supports `scale =` and `offset =` arguments — it transforms on write, inverts on read.
- The user needs to know (calculate) these values beforehand
- The formula to determine `scale` and `offset` can be derived by the min-max formula

$$\text{scale} = \frac{b - a}{\max(x) - \min(x)}, \qquad \text{offset} = \frac{a\,\max(x) - b\,\min(x)}{\max(x) - \min(x)}$$

---


```{r}
#| echo: true

get_scale_offset <- function(x, a = 0, b = 255) {  # x must be a SpatRaster
  library(terra)
  mm    <- minmax(x)                # <3>
  min_x <- mm[1, 1]
  max_x <- mm[2, 1]

  list(
    scale  = (max_x - min_x) / (b - a),                 # <1>
    offset = min_x - a * (max_x - min_x) / (b - a)      # <2>
  )
}
```

1. `writeRaster` *divides* by `scale`: `stored = (x − offset) / scale`
2. `writeRaster` *subtracts* `offset`; for `a = 0` this simplifies to `min_x`
3. Use the global min/max function, not per-cell (e.g. `min()`)


:::{.notes}

- `min(raster)` returns per-cell minimum over bands, NOT the global minimum → always use `minmax()` or `global()`
- terra's writeRaster applies `stored = (x - offset) / scale`; solving for `offset` such that `min_x` maps to `a` gives `offset = min_x - a * scale`


:::


## Scale / offset in practice

```{r}
#| echo: true
library(terra)

elev <- rast(system.file("ex/elev.tif", package = "terra"))
plot(elev, main = paste(minmax(elev), collapse = " – "))
```

---

```{r}
#| echo: true

so <- get_scale_offset(elev)

so

# Note: floating-point rounding may introduce NAs at boundary values
writeRaster(elev, "data-out/INT1U.tif", datatype = "INT1U", overwrite = TRUE,
            scale = so$scale, offset = so$offset)

writeRaster(elev, "data-out/FLT4S.tif", datatype = "FLT4S", overwrite = TRUE)


```

```{r}
file.size("data-out/INT1U.tif") / file.size("data-out/FLT4S.tif")
```

---

Scale and offset are visible in the file metadata — any GDAL-powered tool can read them:

```{bash}
gdal raster info -f json data-out/INT1U.tif | jq '.bands[] | {scale, offset}'
```

:::{.notes}

- Reading `INT1U.tif` back gives original float values — the Byte storage is transparent
- The ratio shows how much smaller the Byte file is compared to Float32 (`FLT4S`)
- In the gdalinfo output, look for `Type=Byte` (stored type) and `Scale`/`Offset` under Band 1

:::


# gdal raster info {background-color="#1269B0"}

Extracting file (meta) data

## gdal raster info

- `gdal raster info` lists information about a raster dataset
- We'll use `elev-lux.tif` — a small elevation model of Luxembourg — as our running example



---

- List all arguments with `--help`:

```{bash}
gdal raster info --help
```


:::{.notes}

- Named args use `--flag` convention; positional arg is `<filename>`
- Default output is JSON (text from GDAL 3.12 when run interactively) — use `-f json` to be explicit

:::


---

- Get all metadata and display output as text

```{bash}
gdal raster info -f text data/raster-deepdive/elev-lux.tif
```

:::{.notes}

- Output includes: driver, dimensions, CRS, origin, pixel size, bands, datatype, NoData value, scale/offset
- This is the quickest way to inspect any raster file without opening R or Python

:::

---

- The metadata output format (`-f`) can either be `text` or to `json`
- Where text is more human readable, `json` is structured and machine readable

```{bash}
gdal raster info -f json data/raster-deepdive/elev-lux.tif
```

---

- JSON in turn can be parsed using [`jq`](https://jqlang.org) — a lightweight command-line JSON processor
  - Navigates JSON with `.key` (field access) and `[]` (array iteration)
  - Install: `conda install -c conda-forge jq`


```{bash}
gdal raster info -f json data/raster-deepdive/elev-lux.tif | jq ".driverLongName"
```

. . .

Extract a single field — `.bands[]` iterates over bands, `.min` selects the field:

```{bash}
gdal raster info -f json data/raster-deepdive/elev-lux.tif | jq ".bands[].min"
```


. . .

Extract multiple fields — pipe `|` into `{min, max}` to select both at once:

```{bash}
gdal raster info -f json data/raster-deepdive/elev-lux.tif | jq '.bands[] | {min, max}'
```


## gdal raster info: terminal histogram

Extract histogram buckets and visualise with [bashplotlib](https://github.com/glamp/bashplotlib):

```{bash}
#| eval: false
gdal raster info --hist -f json data/raster-deepdive/elev-lux.tif |
  jq ".bands[].histogram.buckets[]" | # <1>
  hist -w 1 -p +                      # <2>
```

1. Extract raw bucket counts from the histogram array
2. `hist` from bashplotlib

---


```
 19|  +
 18|  +
 17|  +
 16|  +
 15|  +
 14|  +
 13|  +
 12|  +    +
 11|  +    +
 10| ++ ++ +
  9| ++ ++++                  + +
  8| +++++++  +               + +
  7| ++++++++ ++++ +    ++    + +
  6| ++++++++ ++++ +   +++    + +
  5| +++++++++++++ +  ++++    + ++ +
  4| ++++++++++++++++ ++++++  + ++ ++     +
  3| +++++++++++++++++++++++ ++ ++ ++   + +  + +         +  +
  2| +++++++++++++++++++++++++++++ ++++ + + ++ + +   +   +  + +
  1| ++++++++++++++++++++++++++++++++++ ++++++++ + +++  +++ +++ + +  +   ++
    ----------------------------------------------------------------------
```


:::{.notes}

- Install bashplotlib with `pip install bashplotlib`
- Useful for quick EDA directly in the terminal without opening R/Python

:::


# NoData in rasters {background-color="#1269B0"}

## NoData in rasters

- Raster files have **no explicit null type** — every cell must hold a numeric value
- To mark cells as missing, a two-step approach is used:
  1. Assign a **reserved value** within the datatype's range (e.g. the maximum possible value)
  2. **Label** that value as `NoData` in the file metadata

. . .

→ Any software reading the file interprets that reserved value as missing, not as a real measurement

:::{.notes}

- This is different from R's in-memory `NA`, which is a proper null type
- The reserved value label lives in the metadata, just like scale/offset

:::


## NoData & data types

For **integer types**, the reserved value must fit within the datatype's range:

| Datatype        | Common NoData reserved value |
| :-------------- | :--------------------------- |
| `Byte`          | 255 (max value)              |
| `Int16`         | −32,768 (min value)          |
| `Float32`[^NAN] | NaN or −9999                 |


[^NAN]: Float types support NaN (as part of the IEEE 754 standard). `-9999` sometimes seen in float rasters is a legacy workaround for older software that cannot handle NaN


. . .

**Risk (integer types):** if real data happens to contain the reserved value → those cells silently become `NoData`

:::{.notes}

- For Byte: using 255 as NoData means you only have 0–254 for real data
- NaN is an unambiguous bit pattern in IEEE 754 — it represents "undefined result" (e.g. 0/0) and is never a valid number
- The silent data-loss risk applies only to integer types; float types with NaN are safe

:::


## NoData in terra

Inspect and set the NoData flag:

```{r}
#| echo: true
library(terra)

elev <- rast(system.file("ex/elev.tif", package = "terra"))

NAflag(elev)   # current NoData reserved value
```

---

```{r}
#| echo: true
#| fig-cap: "Note that the maximum elevation values are now NA"
#| classes: .center

# Only use 0 - 254, so that 255 is availalable for "NoData"
so <- get_scale_offset(elev, a = 0, b = 254)

# write with an explicit NoData flag
writeRaster(elev, "data-out/elev_naflag.tif", datatype = "INT1U", overwrite = TRUE,
            scale = so$scale, offset = so$offset, NAflag = 255)

# read back — terra interprets 255 as NA automatically
elev_back <- rast("data-out/elev_naflag.tif")
plot(elev_back, colNA = "black")
```

---

The reserved value and its NoData label are both visible in the metadata:

```{bash}
gdal raster info -f json data-out/elev_naflag.tif | jq ".bands[].noDataValue"
```

:::{.notes}

- `NAflag()` returns the value currently flagged as NoData; `NA` means no flag is set
- On read, terra converts the flagged value back to R's `NA` transparently
- `NAflag` and `datatype` are set together in `writeRaster` — they are two sides of the same coin
- In the gdalinfo output, look for `NoData Value=255` under Band 1

:::


# Categorical rasters {background-color="#1269B0"}


## Categorical rasters

- Some raster data is **categorical**, not continuous: land use, soil type, vegetation class, …
- Raster files cannot store text — categories are encoded as **integers**
- The mapping (integer code → category label) is stored in the file metadata as an **attribute table**

. . .

→ Analogous to R's `factor`: an underlying integer + a levels table

. . .

→ Choose the smallest integer type that covers the number of categories — `Byte` handles up to 255 classes

:::{.notes}

- Most land use / land cover classifications have far fewer than 255 classes → Byte is almost always sufficient
- The attribute table lives in the file metadata, just like scale/offset and NoData flags

:::


## Categorical rasters in terra

```{r}
#| echo: true
elev_cat <- classify(elev, c(0, 200, 350, 550))                                     # <1>
levels(elev_cat) <- data.frame(value = 0:2, label = c("low", "medium", "high"))     # <2>

plot(elev_cat)
```

1. Reclassify values into three categories (`0-200`, `200-350` and `350-550`)
2. Relable the default classes to `low`, `medium` and `high`

---

```{r}
#| echo: true
# write as Byte explicitly (optional, writeRaster makes a reasonable guess)
writeRaster(elev_cat, "data-out/elev_cat.tif", datatype = "INT1U", overwrite = TRUE)
```

[]{.gap}

```{bash}
gdal raster info -f json data-out/elev_cat.tif | jq '.bands[].categories[0:3]'
```
:::{.notes}

- `levels()` sets the attribute table — same function as for R factors
- `cats(elev_cat)` to inspect the table after the fact
- On read, terra restores the levels automatically if the attribute table is embedded in the file

:::


```{r}
#| echo: false

filesize2df <- function(file, comment){
  data.frame(
    comment  = comment,
    size_mb  = round(file.size(file) / 1024^2, 2)
  )
}

bind_print <- function(...){
  rbind(...) |>
    dplyr::mutate(difference = paste(round((size_mb / size_mb[1] - 1) * 100), "%")) |>
    knitr::kable(col.names = c("File", "Size (MB)", "Difference"), align = "lrr")
}
```

# Compression {background-color="#1269B0"}

## Raster file size

- File size = **dimensions × bands × bytes per pixel**
- Example: single SRTM tile, one band, `Int16` (2 bytes/pixel):

$$3601 \times 3601 \times 2\,\text{B} = 25{,}934{,}402\,\text{B} \approx 25.9\,\text{MB}$$

- **Compression** reduces storage by exploiting spatial redundancy in pixel values

:::{.notes}

- Real-world rasters often have large homogeneous regions → high redundancy → compress well
- Redundancy is especially high in categorical rasters (e.g. land cover with few distinct values)

:::


## Lossless vs lossy

|                 | **Lossless**                         | **Lossy**                     |
| :-------------- | :----------------------------------- | :---------------------------- |
| Data preserved? | Perfectly                            | Approximated                  |
| Use case        | Scientific data (DEM, multispectral) | Photographic / visual imagery |
| Examples        | LZW, DEFLATE, ZSTD, PACKBITS         | JPEG2000, WEBP                |

. . .

→ For scientific analysis **always use lossless** — lossy compression permanently alters pixel values

:::{.notes}

- WEBP supports both lossy and lossless modes depending on settings
- Once lossy-compressed, original values cannot be recovered — important for analysis reproducibility
- Much of this section is based on @gandhi2020 and [Paul Ramsey's GeoTIFF compression guide](https://blog.cleverelephant.ca/2015/02/geotiff-compression-for-dummies.html)

:::


## How compression works

Exploit **repeated values** — store each unique value once, track its positions:

|            |                                                                                         |
| :--------- | :-------------------------------------------------------------------------------------- |
| Original   | `100, 101, 102, 100, 100`                                                               |
| Compressed | `100` → positions `[0, 3, 4]` <br> `101` → positions `[1]` <br> `102` → positions `[2]` |

→ 5 values encoded as 3 unique values + a position list

. . .

Works best when data contains **many repeated or near-repeated values** (flat terrain, uniform land cover)

:::{.notes}

- This is a simplified illustration of dictionary / run-length encoding
- LZW and DEFLATE use more sophisticated variants of this principle

:::


## PREDICTOR

LZW, DEFLATE, and ZSTD can use a **predictor** to improve compression further:

→ Store **differences** between consecutive values instead of absolute values

|                |        |       |       |       |       |
| :------------- | :----- | ----- | ----- | ----- | ----- |
| Original       | `100`  | `101` | `102` | `100` | `100` |
| With predictor | `100`  | `+1`  | `+1`  | `−2`  | `0`   |

Small differences compress better than large absolute values

. . .

| Option        | Use for                  |
| :------------ | :----------------------- |
| `PREDICTOR=1` | No predictor (default)   |
| `PREDICTOR=2` | Integer data (e.g. DEMs) |
| `PREDICTOR=3` | Float data               |

:::{.notes}

- Predictor=2 works well for integer rasters with gradual spatial variation
- Predictor=3 for float rasters
- Using the wrong predictor (e.g. =3 on integer data) still works but may not compress as well

:::


## Tiling & compression costs

- By default, GeoTIFF stores data **line by line** (strip layout)
- `TILED=YES` stores data in **256 × 256 pixel blocks**

→ Tiling improves random-access reads and is essential for **Cloud-Optimized GeoTIFF (COG)**

. . .

**Compression is not free:**

- Adds CPU overhead on **write and read**
- Good default for most scientific workflows: **`DEFLATE` + `TILED=YES`**

:::{.notes}

- Strip layout is fast for sequential top-to-bottom reads but slow for spatial queries on large files
- ZSTD generally offers better ratios than DEFLATE but has less universal software support
- COG = Cloud-Optimized GeoTIFF; tiling + overviews enable efficient streaming from cloud storage

:::


## gdal raster convert

[`gdal raster convert`](https://gdal.org/en/stable/programs/gdal_raster_convert.html) converts raster files between formats and applies compression:

```{bash}
gdal raster convert --help
```

Key points:

- Two **positional arguments**: `<input_file>` `<output_file>`
- Compression via `--co KEY=VALUE` (**c**reation **o**ptions)
- Options are format-specific — always check `--help`

:::{.notes}

- Mutually exclusive options are shown as `[[-strict]|[-not_strict]]` in the help output

:::


## gdal raster convert: baseline

```{bash}
#| eval: false
gdal raster convert data/raster-deepdive/dhm25_grid_raster.asc data-out/dhm25.tif
```

```{r}
#| echo: false
size0 <- filesize2df("data/raster-deepdive/dhm25_grid_raster.asc", "Original (ASCII)")
size1 <- filesize2df("data-out/dhm25.tif", "GeoTIFF, no compression")
bind_print(size0, size1)
```


## gdal raster convert: + DEFLATE

```{bash}
#| eval: false
gdal raster convert data/raster-deepdive/dhm25_grid_raster.asc data-out/dhm25_2.tif \
  --co COMPRESS=DEFLATE
```

```{r}
#| echo: false
size2 <- filesize2df("data-out/dhm25_2.tif", "`COMPRESS=DEFLATE`")
bind_print(size0, size1, size2)
```


## gdal raster convert: + TILED

```{bash}
#| eval: false
gdal raster convert data/raster-deepdive/dhm25_grid_raster.asc data-out/dhm25_3.tif \
  --co COMPRESS=DEFLATE \
  --co TILED=YES
```

```{r}
#| echo: false
size3 <- filesize2df("data-out/dhm25_3.tif", "`TILED=YES`")
bind_print(size0, size1, size2, size3)
```


## gdal raster convert: + PREDICTOR

```{bash}
#| eval: false
gdal raster convert data/raster-deepdive/dhm25_grid_raster.asc data-out/dhm25_4.tif \
  --co COMPRESS=DEFLATE \
  --co TILED=YES \
  --co PREDICTOR=3
```

```{r}
#| echo: false
size4 <- filesize2df("data-out/dhm25_4.tif", "`PREDICTOR=3`")
bind_print(size0, size1, size2, size3, size4)
```

:::{.notes}

- `PREDICTOR=3` for floating-point data; use `PREDICTOR=2` for integer data
- Each row in the table adds one `-co` flag — the cumulative effect shows how the options interact

:::


# gdal raster reproject {background-color="#1269B0"}


## gdal raster info: get CRS

To reproject a raster, we need to know the source CRS. Is this information available?

```{bash}
gdal raster info -f json data/raster-deepdive/dhm25_grid_raster.asc | \
  jq ".coordinateSystem"
```



→ No CRS — consult the data provider: DHM25 uses `LV03/LN02` = [`EPSG:21781`](https://epsg.io/21781)


---

Confirm via corner coordinates:

```{bash}
gdal raster info -f json data/raster-deepdive/dhm25_grid_raster.asc  | \
  jq ".cornerCoordinates | {upperLeft, lowerRight}"
```

:::{.notes}

- Large x/y values (e.g. 480000, 70000) are characteristic of Swiss LV03 coordinates → confirms EPSG:21781
- If the CRS field is null, the corner coordinates often reveal the projection — know your coordinate ranges

:::


## gdal raster reproject

- [`gdal raster reproject`](https://gdal.org/en/stable/programs/gdal_raster_reproject.html) reprojects and transforms raster data
- Two positional arguments: `<input>` and `<output>`

```{bash}
gdal raster reproject --help
```

:::{.notes}

- gdal raster reproject can also resample — reprojection is its primary use case
- Key flags: `--src-crs` (source CRS), `--dst-crs` (target CRS), `--resampling` (resampling method)

:::


## gdal raster reproject: reproject

DHM25 has no embedded CRS → specify source and target explicitly:

```{bash}
#| eval: false
gdal raster reproject \
  --src-crs EPSG:21781 \
  --dst-crs EPSG:2056 \
  data/raster-deepdive/dhm25_grid_raster.asc \
  data-out/dhm25.tif
```

```
0...10...20...30...40...50...60...70...80...90...100 - done.
```

:::{.notes}

- EPSG:21781 = old Swiss coordinate system (LV03 / CH1903)
- EPSG:2056 = new Swiss coordinate system (LV95 / CH1903+)
- Output as GeoTIFF — unlike ASC, it supports embedded CRS metadata

:::


## gdal raster reproject: verify

Compare before and after — coordinates should shift from LV03 to LV95 range:

```{bash}
gdal raster info -f json data-out/dhm25.tif | \
  jq '.stac."proj:epsg"'
```

. . .

```{bash}
gdal raster info -f json data-out/dhm25.tif | \
  jq ".cornerCoordinates | {upperLeft, lowerRight}"
```


:::{.notes}

- Corner coordinates in LV95: easting ~2.6M, northing ~1.2M (vs ~480k, ~70k in LV03)
- `.coordinateSystem.wkt` gives the full CRS definition — look for `EPSG:2056` to confirm
- Comparing the corner coordinates to the earlier ASC output is the most intuitive check

:::


# Bringing it all together {background-color="#1269B0"}

## Three tools, one goal

All three strategies reduce on-disk file size — but they work at **different levels**:

| Tool | Targets | Mechanism |
|:-----|:--------|:----------|
| **Datatype** | Per-pixel cost | 1 byte (`Byte`) vs 4 bytes (`Float32`) |
| **Scale / offset** | Value range | Rescale floats to fit a smaller integer type |
| **Compression** | Spatial redundancy | Exploit repeated values across pixels |

. . .

→ They are **complementary** — apply all three for maximum effect

→ Order matters: choose datatype first, then compress

:::{.notes}

- Getting the datatype wrong means you compress a larger file to begin with
- Scale/offset is what makes integer types viable for continuous data — without it you'd lose the value range
- Compression then removes the spatial redundancy that remains regardless of datatype

:::


## Combined example {visibility="hidden"}

Rows 1 & 2 reuse the `dhm25` variants from the compression section; rows 3 & 4 add scale/offset:

```{r}
#| echo: true
library(terra)

dhm <- rast("data/raster-deepdive/dhm25_grid_raster.asc")
so  <- get_scale_offset(dhm)

# reuses dhm25.tif (baseline) and dhm25_4.tif (best compression) from earlier

# 3. Byte + scale/offset, no compression
writeRaster(dhm, "data-out/dhm25_byte.tif",
            datatype = "INT1U", overwrite = TRUE,
            scale = so$scale, offset = so$offset)

# 4. Byte + scale/offset + compression
writeRaster(dhm, "data-out/dhm25_byte_comp.tif",
            datatype = "INT1U", overwrite = TRUE,
            scale = so$scale, offset = so$offset,
            gdal = c("COMPRESS=DEFLATE", "TILED=YES"))
```

## {visibility="hidden"}

```{r}
#| echo: false
s0 <- filesize2df("data-out/dhm25.tif",           "GeoTIFF, no compression (baseline)")
s1 <- filesize2df("data-out/dhm25_4.tif",         "DEFLATE + TILED + PREDICTOR")
s2 <- filesize2df("data-out/dhm25_byte.tif",      "Byte + scale/offset")
s3 <- filesize2df("data-out/dhm25_byte_comp.tif", "Byte + scale/offset + DEFLATE + TILED")
bind_print(s0, s1, s2, s3)
```

:::{.notes}

- Row 2 shows what compression alone buys
- Row 3 shows what datatype reduction alone buys (scale/offset makes it possible)
- Row 4 shows the multiplicative effect of combining both strategies
- The two strategies are not redundant — each contributes independently

:::


## Practical checklist

When writing a raster to disk:

1. **What is my value range?** → choose the smallest datatype that fits
2. **Can I rescale to fit an integer type?** → use `scale` / `offset` in `writeRaster`
3. **Are there missing cells?** → set `NAflag` (integer types) or use NaN (float types)
4. **Apply lossless compression:** `COMPRESS=DEFLATE` + `TILED=YES`
   - Add `PREDICTOR=2` for integer data, `PREDICTOR=3` for float data

. . .

**Default recommendation for most scientific rasters:**

```
INT1U / Int16  +  scale/offset  +  DEFLATE  +  TILED=YES  +  PREDICTOR=2
```

:::{.notes}

- For data that genuinely needs float precision (e.g. model outputs with many significant digits), skip the scale/offset step and use FLT4S + DEFLATE + TILED + PREDICTOR=3
- ZSTD can replace DEFLATE for better ratios where broad software compatibility is less critical

:::


## References
