---
title: Topological relations 
---

# Named topological relations



---

[]{.lts .content-hidden unless-profile="book"}


:::{.incremental}

- Topological relations describe the spatial relationships between objects. 
- They are also called *binary topological relationships* or *binary predicates*
- They are logical statements (`TRUE`/`FALSE`) about the spatial relationships between two objects
- The two objects are defined by ordered sets of points (typically forming points, lines and polygons) in two or more dimensions [@egenhofer1990]. 

:::

## Usage

- Topological relations can be used to subset or join spatial data
- For example:
  - Subsetting: *Return all rivers that flow through the canton of Zurich*
  - Joining: *For every train station, give me the name of the municipality it lies within*

## Common predicates in `sf` {#sec-named-topological-relations}

- Most GIS software offers common topological relations as functions
- `sf` provides many: `st_intersects()`, `st_disjoint()`, `st_touches()`, `st_crosses()`, ...
- Each works slightly differently and fits different contexts
- Examples:
  - `st_covers(x, y)` → `TRUE` if no points of `x` are outside `y`
  - `st_touches(x, y)` → `TRUE` if geometries share boundary points but interiors don't intersect
  
## Key predicates visualized

```{r}
#| echo: false
#| label: fig-topological-overview
#| fig-width: 13
#| fig-cap: "Key topological relations between two polygons (A in purple, B in green)"
#| fig-subcap:
#|   - "`st_disjoint`: no shared points"
#|   - "`st_touches`: shared boundary only"
#|   - "`st_intersects`: any shared points (here with overlap)"
#|   - "`st_contains`: B entirely within A"
#| layout-ncol: 2
#| out-width: "100%"

library(ggplot2)
library(sf)

rect_sfc <- function(xmin, ymin, xmax, ymax) {
  st_sfc(st_polygon(list(cbind(
    c(xmin, xmax, xmax, xmin, xmin),
    c(ymin, ymin, ymax, ymax, ymin)
  ))))
}

plot_pair <- function(a, b) {
  ca <- st_sf(geometry = a, label = "A")
  cb <- st_sf(geometry = b, label = "B")
  ggplot() +
    geom_sf(data = ca, fill = "#beaed4", alpha = 0.6) +
    geom_sf(data = cb, fill = "#7fc97f", alpha = 0.6) +
    geom_sf_text(data = ca, aes(label = label), size = 5) +
    geom_sf_text(data = cb, aes(label = label), size = 5) +
    theme_void()
}

# Disjoint
plot_pair(rect_sfc(0, 0, 1, 1), rect_sfc(1.5, 0, 2.5, 1))

# Touches
plot_pair(rect_sfc(0, 0, 1, 1), rect_sfc(1, 0, 2, 1))

# Intersects (overlapping)
plot_pair(rect_sfc(0, 0, 1.5, 1), rect_sfc(1, 0, 2.5, 1))

# Contains
plot_pair(rect_sfc(0, 0, 3, 2), rect_sfc(0.5, 0.3, 2.5, 1.7))
```

## Symmetry and order

- Some relations are **symmetrical** (order doesn't matter)
  - `st_touches(x, y)` = `st_touches(y, x)`
- Others are **asymmetrical** (order matters!)
  - `st_contains(x, y)` ≠ `st_contains(y, x)`
- Some require extra arguments
  - `st_is_within_distance()` needs a `dist` argument
- Full list: `?geos_binary_pred`

## Contains, covers, within

- Easily confused — here's the distinction:
  - `st_contains(x, y)`: is `y` inside `x`, with ≥1 point in the **interior**?
  - `st_covers(x, y)`: is every point of `y` inside **or on the boundary** of `x`?
  - `st_within(x, y)`: inverse of `st_contains` — is `x` inside `y`?
- Key difference: a point *on the boundary* is **covered** but not **contained**
- **Tip**: for point-in-polygon queries, prefer `st_covers` — `st_contains` misses points exactly on the boundary

```{r}
#| collapse: true
polygon <- st_sfc(st_polygon(list(cbind(c(0,1,1,0,0), c(0,0,1,1,0)))))
point_inside <- st_sfc(st_point(c(0.5, 0.5)))
point_on_boundary <- st_sfc(st_point(c(0, 0.5)))

st_contains(polygon, point_inside, sparse = FALSE)      # TRUE
st_contains(polygon, point_on_boundary, sparse = FALSE)  # FALSE
st_covers(polygon, point_on_boundary, sparse = FALSE)    # TRUE
```

## Predicates vs. operations

:::{.callout-important}
## Don't confuse predicates with operations!

- **Predicate** (e.g. `st_intersects`): *"Do these geometries intersect?"* → returns `TRUE`/`FALSE`
- **Operation** (e.g. `st_intersection`): *"Give me the geometry where they overlap"* → returns a new geometry

The same naming pattern applies to other pairs: `st_difference` (operation) has no corresponding predicate, while `st_touches` (predicate) has no corresponding operation.
:::

## Predicate return types

- `st_intersects(x, y)` does **not** return a simple `TRUE`/`FALSE` vector
- Returns a *sparse geometry binary predicate* (`sgbp`) — a list of **matching indices**

```{r}
#| collapse: true
a <- rect_sfc(0, 0, 1, 1)          # one polygon
b <- c(
  rect_sfc(0.5, 0, 1.5, 1),        # overlaps a
  rect_sfc(2, 0, 3, 1)             # disjoint from a
)

st_intersects(a, b)
```

- `lengths()` counts the number of matches per feature
- Use `sparse = FALSE` to get a dense logical matrix instead

```{r}
#| collapse: true
st_intersects(a, b, sparse = FALSE)
```

## Applying predicates: subsetting {.center}

Now that we know how predicates work, let's use them to **subset** and **join** spatial data.

## Spatial subsetting

- Predicates can be used to **subset** one dataset based on another
- Example data: playgrounds, public transport stops, and Stadtkreise in Zurich

```{r}
library(sf)
library(readr)

# source: https://www.stadt-zuerich.ch/geodaten/
playgrounds <- read_sf("data/week4-exercises/playgrounds.gpkg")
publictransport <- read_sf("data/week4-exercises/public_transport.gpkg")
kreise <- read_sf("data/week4-exercises/stadtkreise-zh.gpkg")
```

- Default predicate: `st_intersects`
- Example: *which playgrounds lie within Stadtkreis 1?*

```{r}
#| collapse: true
kreis1 <- kreise[kreise$STADTKREIS == "Kreis 1", ]

playgrounds_k1 <- playgrounds[kreis1, ]  # st_intersects is the default

nrow(playgrounds_k1)
```

## Distance-based subsetting

- Not limited to topological predicates — can also use distance-based ones
- Example: *which playgrounds are within 100m of a public transport stop?*

```{r}
# Using the shorthand notation
playgrounds_close <- playgrounds[publictransport,,op = st_is_within_distance, dist = 100]
```

Same thing, more readable with `st_filter()`:

```{r}
playgrounds_close <- st_filter(playgrounds, publictransport, .predicate = st_is_within_distance, dist = 100)
```

---

```{r}
#| echo: false
#| label: fig-playground-publictransport
#| fig-cap: "Note that the playgrounds within 100m of public transport (red dots) are a *subset* of all the playgrounds"

library(tmap)  

tm_shape(publictransport) + tm_dots(shape = 2, col = "darkblue") +
tm_shape(playgrounds) + tm_dots(shape = 16, fill = "darkgreen", size = .5)  +
  tm_shape(playgrounds_close) + tm_dots(shape = 16, fill = "red", size = .5) +
  tm_add_legend(fill = c("darkgreen","red", "darkblue"), col = c("darkgreen", "red", "darkblue"), shape = c(16,16, 2),labels = c("Playgrounds", "Playgrounds within 100m", "Public Transport")) 


```

## From subsetting to joining {.center}

Subsetting *filters* features — what if we want to *transfer attributes* instead?

## Spatial joins

- **Joins** transfer attributes from one dataset to another based on spatial relationship
- `st_join(x, y)`: for each feature in `x`, find matching features in `y` and attach their columns
- Example: add the name of the *nearest* public transport stop to each playground

:::{.notes}
`st_nearest_feature` is not strictly a binary predicate, but is very useful for spatial joins. To make the example clearer, we first discard all unnecessary columns from the datasets.
:::

---

```{r}

publictransport <- publictransport[,"CHSTNAME"]

playgrounds <- playgrounds[, "name"]


playgrounds_join <- st_join(
  playgrounds, 
  publictransport, 
  join = st_nearest_feature
  )
```


```{r}
playgrounds_join
```

## Join cardinality

- `playgrounds_join` has the same rows as `playgrounds` + extra column `CHSTNAME`
  - Why? Each playground has exactly **one** nearest station
- But row count can change with other join predicates!
  - *Within 100m*: a playground may match **zero or multiple** stops

---

```{r}
#| collapse: true
playgrounds_join2 <- st_join(
  playgrounds, 
  publictransport, 
  join = st_is_within_distance, 
  dist = 100
  )

nrow(playgrounds)

nrow(playgrounds_join2)

```

---

```{r}
playgrounds_join2
```

## Join order

- Order matters (just like regular joins)
- `st_join` = **left join** by default
- Result keeps the **geometry of the left dataset**

---

```{r}
kreise <- kreise[,"STADTKREIS"]

publictransport_join <- st_join(publictransport, kreise)
```

```{r}
#| echo: false
tm_shape(publictransport_join) + tm_dots(fill = "STADTKREIS")
```

---

- Reverse order → each `stadtkreis` gets **duplicated** for every intersecting point

```{r}
#| collapse: true

kreise_join <- st_join(kreise, publictransport)

nrow(kreise)

nrow(kreise_join)
```


```{r}
#| echo: false
#| label: fig-publictransport-kreise
#| fig-cap: "Note that all the duplicate stadtkreise overlap each other, so when you visualize the data, the issue is not noticeable"
#| fig-subcap: 
#|   - "Dataset `kreise` with 12 features"
#|   - "Dataset `kreise_join` with 477 features"
#| layout-nrow: 1
tm_shape(kreise) + tm_polygons("STADTKREIS") + tm_layout(legend.show = FALSE)

tm_shape(kreise) + tm_polygons("STADTKREIS") + tm_layout(legend.show = FALSE)

```

## Summary

| Task | Function | Key arguments |
|---|---|---|
| Test a relationship | `st_intersects()`, `st_covers()`, ... | `sparse = FALSE` for logical matrix |
| Subset by predicate | `x[y, ]` or `st_filter()` | `op` / `.predicate`, `dist` |
| Join by predicate | `st_join(x, y)` | `join`, `left = TRUE` |

- Default predicate is always `st_intersects`
- Prefer `st_covers` over `st_contains` for point-in-polygon
- Join order determines geometry and potential row duplication

# Custom topological relations

## A motivating problem

- Consider a 3×3 chessboard (@fig-chessboard)
- Which fields share a **full edge** with the origin? (like a rook's move)
- `st_touches` won't work — it also matches **diagonal** neighbours (shared point)
- We need a way to express: *"shared boundary must be a line, not just a point"*

```{r}
#| echo: false
#| out-width: "80%"
#| label: fig-chessboard
#| fig-cap: A 3x3 chessboard with a rook in the center field (origin). Which fields can the rook reach, if the constraint is that the destination field need to share an edge with the origin?
xy2sfc = function(x, y) st_sfc(st_polygon(list(cbind(x, y))))
x = xy2sfc(x = c(0, 0, 1, 1, 0), y = c(0, 1, 1, 0, 0))

grid = st_make_grid(x, n = 3)

grid_dest <- grid[c(1:4, 6:9)]
grid_orig <- grid[5]

chessplot <- tm_shape(grid_dest) + tm_polygons(fill = "#7fc97f") +
  tm_shape(grid_orig) + tm_polygons(fill = "#beaed4") +
  tm_shape((x)) +
  tm_dots(shape =tmap_icons("images/rook.png"),size = 1.2) +
    tm_layout(frame = FALSE) +
  tm_add_legend(
    labels = c("Origin","Destination"),
    fill = c("#beaed4","#7fc97f"),
    type = "polygons"
    )

chessplot
```


```{r}
#| echo: false
#| output: false

st_write(grid_dest, "data/week4-exercises/chessboard.gpkg","grid_dest", append = FALSE)

st_write(grid_orig, "data/week4-exercises/chessboard.gpkg","grid_orig", append = FALSE)


```

## The DE-9IM

- **DE-9IM** (*Dimensionally Extended nine-Intersection Model*) gives fine-grained control
- DE-9IM is the formal model **behind** all named predicates
- Describes the relationship as a **3×3 matrix**: interior, boundary, exterior of each geometry
- @tbl-de9im shows the analysis for two overlapping polygons

---


```{r}
#| echo: false
#| label: tbl-de9im
#| tbl-cap: "Image source: M. W. Taves [commons.wikimedia.org](https://commons.wikimedia.org/wiki/File:DE9IM_II.svg)"


library(dplyr)
library(tidyr)


perspec <- c("Interior", "Boundary", "Exterior")
perspec2 <- c("I","B","E")

df <-cbind(rep(perspec2, each = 3), rep(perspec2, 3)) |>
  apply(1, \(x)paste(x,collapse = "")) |>
  (\(x)file.path("images","DE9IM",paste0("DE9IM_",x,".svg.png")))() |>
  matrix(ncol = 3, byrow = TRUE) |>
  as.data.frame()

colnames(df) <- perspec
rownames(df) <- perspec

library(gt)
df |>
  gt::gt(rownames_to_stub = TRUE) |>
  fmt_image(columns = c(Interior,Boundary,Exterior),height = "4em")

```

---

- Reading @tbl-de9im:
  - Interior ∩ Interior → area → `2`
  - Interior ∩ Boundary → line → `1`
  - Boundary ∩ Boundary → points → `0`
- Flatten all nine cells into a string → `212101212`
- The power: **specify the pattern you want** and match geometries against it

:::{.notes}
The DE-9IM string is read row by row from the 3×3 matrix. The encoding is: F = empty, 0 = point, 1 = line, 2 = area. This concept is especially useful from the "opposite" perspective: we describe the DE-9IM pattern we are looking for and use `st_relate` to find matching pairs.
:::

## Predicates as patterns

- Every named predicate = one or more DE-9IM patterns

| Predicate | DE-9IM pattern(s) |
|---|---|
| `st_intersects` | `T********`, `*T*******`, `***T*****`, `****T****` (any non-empty intersection) |
| `st_touches` | `FT*******`, `F**T*****`, `F***T****` |
| `st_contains` | `T*****FF*` |
| `st_within` | `T*F**F***` |

- `st_contains` and `st_within` are mirror images (swap rows ↔ columns)
- `*` = wildcard, matches `F`, `0`, `1`, or `2`

## Back to the rook

- Now we can express our rook requirement as a DE-9IM pattern
- Shared boundary = **line**, interiors **don't** intersect
- DE-9IM encoding (@tbl-rook):
  - Interior ∩ Interior = `F` (no overlap)
  - Boundary ∩ Boundary = `1` (shared edge = line)
  - Everything else = `*` (don't care)

```{r}
#| label: tbl-rook
#| tbl-cap: "The DE-9IM for the *rooks* case. This can be flattend into the string `F***1****`"
#| echo: false


df_rook <- stringr::str_split("F***1****","")[[1]] |> 
  matrix(ncol = 3, byrow = TRUE,dimnames = list(perspec,perspec)) |> 
  as.data.frame()

gt(df_rook,rownames_to_stub = TRUE)
```

## `st_relate`: explore

- Without `pattern`: returns the **actual DE-9IM string** for each pair
- Useful for exploration → inspect strings, then decide which pattern to match

```{r}
st_relate(grid_orig, grid_dest)
```

## `st_relate`: match

- Pass a `pattern` argument → returns `TRUE`/`FALSE` (sparse matrix, like named predicates)

```{r}
st_relate(grid_orig, grid_dest, pattern = "F***1****")
```

---

Create a custom `st_rook` function and use it like any named predicate:

```{r}
st_rook <- \(x, y) st_relate(x, y, pattern = "F***1****")

grid_rook <- grid_dest[grid_orig, , op = st_rook] |> 
  st_sample(1000, type = "hexagonal",by_polygon = TRUE)
```


```{r}
#| echo: false
#| out-width: "80%"
#| label: fig-chessboard-rook
#| fig-cap: The chessboard situation with the potential fields for the rook highlighted with a red outline


chessplot +
  tm_shape(grid_rook) + tm_dots() +
  tm_add_legend(
    labels = c("Rooks case"),
    type = "dots"
    )
```


# References