---
title: "DuckDB"
---



- Open-source, column-oriented **RDBMS** (relational database management system)
- High performance on complex queries, even on large datasets
- Flexible [extension](https://duckdb.org/docs/stable/extensions/overview) system — [spatial](https://duckdb.org/docs/stable/extensions/spatial/overview.html) is a core extension

:::{.notes}
For a great introduction by DuckDB's co-creator, watch [Hannes Mühleisen — DuckDB, an in-process analytical DBMS](https://www.youtube.com/watch?v=Z-6SnP6yzgo) (NY Open Statistical Programming Meetup, 2022).
:::

## Why DuckDB for spatial data?

- R/`sf` loads entire datasets into memory → bottleneck with large data
- DuckDB can **stream from disk** — processes data exceeding available RAM
- SQL: declarative language — describe *what* you want, not *how*
- Spatial extension uses familiar function names (`st_intersects`, `st_area`, ...)
- **Spatial indexing** (R-Tree) makes large-scale spatial queries fast

## Reinventing DBMS?

- `dplyr`, `data.table`, `arrow` do relational transformations (select, join, aggregate, ...)
- But compared to a real DBMS, they lack:
  - **Holistic query optimization** across a pipeline
  - **Out-of-memory computation**
  - **Parallelism and pipelining**
- DuckDB brings 50+ years of DBMS research into your R/Python session

:::{.notes}
Hannes Mühleisen (DuckDB co-creator) makes this point in his [talk](https://www.youtube.com/watch?v=Z-6SnP6yzgo): "The data science community is reinventing database management systems, but rather poorly." If you build a dplyr or data.table pipeline, there is no optimization step that looks at the entire pipeline — it just materializes after each step, leading to potentially huge intermediate results.
:::

## OLAP vs. OLTP

| | **OLAP** (Analytical) | **OLTP** (Transactional) |
|---|---|---|
| Workload | Read-mostly | Many small writes |
| Queries | Complex, scan large parts | Simple, touch individual rows |
| Updates | Bulk appends | Frequent row-level updates |

- DuckDB = **OLAP**

## In-process vs. standalone

- **In-process**: runs inside your application (no separate server)
  - DuckDB, SQLite
- **Standalone**: separate server process, client connects over network
  - PostgreSQL, MySQL


```{r}
#| echo: false
#| label: tbl-duckdb
#| tbl-cap: DuckDB fills a niche that no previous software has filled yet

library(gt)

df2 <- c("sqlite","duckdb","postgres","clickhouse") |> 
  (\(x)file.path("images","logos",paste0(x, ".png")))() |> 
  matrix(ncol = 2, byrow = TRUE, dimnames = list(c("In-Process","Stand-Alone"),c("OLTP","OLAP"))) |> 
  data.frame()
  
df2 |> 
  gt(rownames_to_stub = TRUE) |> 
  fmt_image(columns = c(OLTP, OLAP), width = "10em")

```


## Let's put DuckDB to work {.center}

## DuckDB in practice {#sec-duckdb-practice}

- Download `wald-kantone.duckdb` from Moodle (forest + canton boundaries)
- Install [DuckDB CLI](https://duckdb.org/docs/installation/?version=stable&environment=cli) + [R package](https://duckdb.org/docs/installation/?version=stable&environment=r)
- Install [DBeaver Community](https://dbeaver.io/download/)
- Connect to the database in DBeaver
- Install and load the spatial extension:
   
```{.sql}
INSTALL spatial;
LOAD spatial;
```

---

- Verify both tables are present and explore them:

```{.sql}
SHOW TABLES;
SELECT * FROM wald;
SELECT * FROM kantone;
```

- Create an R-Tree spatial index for both tables:

```{.sql}
CREATE INDEX kantone_idx ON kantone USING RTREE (geom);
CREATE INDEX wald_idx ON wald USING RTREE (geom);
```

---

## Why spatial indices?

- Without an index: `WHERE st_intersects(a, b)` checks **every** pair → slow
- **R-Tree** index organizes geometries by bounding boxes
- Eliminates non-overlapping pairs cheaply → only candidates get the expensive exact test
- Reduces comparisons dramatically (e.g. millions → thousands)

:::{.notes}
R-Trees are the standard spatial index in PostGIS, SpatiaLite, DuckDB, and also power `sf`'s spatial predicates internally.
:::

---

## Building an analysis step-by-step

- Indices speed up queries — but we still need to **compose** the analysis
- Goal: compute **forest share per canton** — intersect forest polygons with canton boundaries, then aggregate areas
- Strategy: break it into small, reusable pieces using SQL `VIEW`s

## SQL `VIEW`s

- A `VIEW` = a **named SQL query** (virtual table)
- Re-executed every time you access it
- Lets us build complex analyses **step-by-step**

:::{.notes}
Unlike materialized tables, VIEWs don't store data on disk — no extra space, always up-to-date. Trade-off: re-computed on every access. We will use VIEWs to build a forest-per-canton analysis in SQL.
:::

- Our first `VIEW`: a subset of the forest dataset (for faster iteration)

---

- Limit to 1'000 rows and store as a `VIEW`:

```{.sql}
CREATE VIEW wald2 AS            -- <1>
SELECT * FROM wald LIMIT 1000;  -- <2>
```

1. Prepend `CREATE VIEW name AS` to any `SELECT`...
2. ... to store it as a reusable virtual table

- Query it like a regular table:

```{.sql}
SELECT * FROM wald2;
```

---

## Spatial intersection in SQL

- Step 1 of the pipeline: intersect forest polygons with canton boundaries

```{.sql}
SELECT 
  name, 
  st_intersection(w.geom, k.geom),  -- <2>
FROM wald2 w, kantone k;            -- <1>
```

1. `w` and `k` are aliases...
2. ... used in the intersection

```{.txt}
┌──────────────────────┬───────────────────────────────────────────────────────┐
│         name         │            st_intersection(w.geom, k.geom)            │
│       varchar        │                       geometry                        │
├──────────────────────┼───────────────────────────────────────────────────────┤
│ Genève               │ POLYGON EMPTY                                         │
│ Genève               │ POLYGON EMPTY                                         │
│ Genève               │ POLYGON EMPTY                                         │
│ Genève               │ POLYGON EMPTY                                         │
│ Genève               │ POLYGON EMPTY                                         │
│   ·                  │       ·                                               │
│   ·                  │       ·                                               │
│   ·                  │       ·                                               │
│ Appenzell Innerrho…  │ POLYGON Z ((2752421.3200000883 1239807.9399965794 9…  │
│ Appenzell Innerrho…  │ POLYGON EMPTY                                         │
│ Appenzell Innerrho…  │ POLYGON EMPTY                                         │
│ Appenzell Innerrho…  │ POLYGON EMPTY                                         │
│ Appenzell Innerrho…  │ POLYGON EMPTY                                         │
├──────────────────────┴───────────────────────────────────────────────────────┤
│ 26000 rows (10 shown)                                              2 columns │
└──────────────────────────────────────────────────────────────────────────────┘
Run Time (s): real 1.292 user 1.293885 sys 0.005016


```


---

- Problem: **cross join** — every forest polygon × every canton (1'000 × 26 = 26'000 pairs)
- Solution: add `WHERE` to filter early:

```{.sql}
SELECT
  name,
  st_intersection(w.geom, k.geom),
FROM wald2 w, kantone k
WHERE st_intersects(w.geom, k.geom); -- <1>
```

1. Only compute the intersection for pairs that actually overlap (R-Tree!)


```{.txt}
┌──────────────────────┬───────────────────────────────────────────────────────┐
│         name         │            st_intersection(w.geom, k.geom)            │
│       varchar        │                       geometry                        │
├──────────────────────┼───────────────────────────────────────────────────────┤
│ Appenzell Innerrho…  │ POLYGON Z ((2750577.8690000786 1236499.1299968604 1…  │
│ St. Gallen           │ POLYGON Z ((2733301.3710000697 1235901.1909968755 1…  │
│ St. Gallen           │ POLYGON Z ((2730828.668000072 1236533.6979968112 10…  │
│ Appenzell Innerrho…  │ POLYGON Z ((2751117.7790000793 1236511.4949968604 1…  │
│ Zürich               │ POLYGON Z ((2712200.8850000626 1236517.2959967866 6…  │
│   ·                  │                           ·                           │
│   ·                  │                           ·                           │
│   ·                  │                           ·                           │
│ Appenzell Innerrho…  │ POLYGON Z ((2752421.3200000883 1239807.9399965794 9…  │
│ Zürich               │ POLYGON Z ((2684156.710000051 1239693.0049964704 51…  │
│ St. Gallen           │ POLYGON Z ((2734307.2300000787 1239757.5959965463 1…  │
│ Appenzell Ausserrh…  │ POLYGON Z ((2735958.5650000805 1239767.1879965463 1…  │
│ Appenzell Ausserrh…  │ POLYGON Z ((2738798.7040000805 1239749.8799965512 8…  │
├──────────────────────┴───────────────────────────────────────────────────────┤
│ 1027 rows (10 shown)                                               2 columns │
└──────────────────────────────────────────────────────────────────────────────┘
Run Time (s): real 0.818 user 0.823316 sys 0.006000

```


:::{.note}
## `st_intersects` vs `st_intersection`
This query uses both: `st_intersects` in `WHERE` = **predicate** (true/false filter), `st_intersection` in `SELECT` = **operation** (computes geometry). Same pattern as in R/`sf`!
:::


---

- We need the **area**, not the geometry itself:


```{.sql}
SELECT 
  name, 
  st_area(st_intersection(w.geom, k.geom)) as wald_area, -- <1>
FROM wald2 w, kantone k
WHERE st_intersects(w.geom, k.geom);
```

1. `st_area` calculates the area of the intersection


```{.txt}
┌────────────────────────┬────────────────────┐
│          name          │     wald_area      │
│        varchar         │       double       │
├────────────────────────┼────────────────────┤
│ Appenzell Innerrhoden  │  3223.381481670876 │
│ St. Gallen             │ 490040.98548061884 │
│ St. Gallen             │ 1515.7608568453097 │
│ Appenzell Innerrhoden  │ 5697.7576364166425 │
│ Zürich                 │  9248.165822466484 │
│   ·                    │          ·         │
│   ·                    │          ·         │
│   ·                    │          ·         │
│ Appenzell Innerrhoden  │ 31971.908096492414 │
│ Zürich                 │ 11339.135541535818 │
│ St. Gallen             │ 11570.958247649043 │
│ Appenzell Ausserrhoden │ 3659.7222327571385 │
│ Appenzell Ausserrhoden │ 1371.8734574811795 │
├────────────────────────┴────────────────────┤
│ 1027 rows (10 shown)              2 columns │
└─────────────────────────────────────────────┘
Run Time (s): real 0.810 user 0.811628 sys 0.018473

```

---

- Save this as a `VIEW` before aggregating:

```{.sql}
CREATE VIEW wald_kantone AS            -- <1>
SELECT 
  name, 
  st_area(st_intersection(w.geom, k.geom)) AS wald_area,
FROM wald2 w, kantone k
WHERE st_intersects(w.geom, k.geom);
```

1. This creates a `VIEW` from the preceding query

---

- Query the `VIEW` like a table, then aggregate with `GROUP BY`:

```{.sql}
SELECT 
  name,                         -- <3>
  sum(wald_area) as wald_area   -- <2>
FROM wald_kantone
GROUP BY name;                  -- <1>
``` 

1. If we use `GROUP BY` in a SQL query...
2. ... we need to wrap all columns with aggregate function...
3. ... except for the columns that we use for grouping

---

- Save the aggregation as a `VIEW`:

```{.sql}
CREATE VIEW wald_kanton_grp AS
SELECT 
  name, 
  sum(wald_area) as wald_area
FROM wald_kantone
GROUP BY name;
```

---

- Join with `kantone` to get the total canton area and compute the fraction:

```{.sql}
SELECT 
	kantone.name,                          -- <3>
	wald_area/area as waldanteil,          -- <4>
FROM wald_kanton_grp 
LEFT JOIN kantone                        -- <1>
ON wald_kanton_grp.name=kantone.name;    -- <2>
```

1. `LEFT JOIN` appends columns from another table...
2. ... matched by the `ON` condition
3. We only need the canton name...
4. ... and the fraction `wald_area / area`

---

- Save as a final `VIEW`, ordered by forest share:

```{.sql}
CREATE VIEW kanton_frac AS
SELECT 
	kantone.name,                 
	wald_area/area as waldanteil, 
FROM wald_kanton_grp 
LEFT JOIN kantone 
ON wald_kanton_grp.name=kantone.name
ORDER BY waldanteil DESC;           -- <1>
```

1. We can `ORDER BY` to show us the highest values first

---

## Scaling up

- So far: only 1'000 forest features → incomplete results
- Since we used `VIEW`s, switching to the full dataset is trivial:

```{.sql}
CREATE OR REPLACE VIEW wald2 AS
SELECT * FROM wald;
```

- Every downstream `VIEW` now automatically uses the full data:

```{.sql}
SELECT * FROM kanton_frac;
```

:::{.notes}
`CREATE OR REPLACE VIEW` is needed because `wald2` already exists. Key advantage of VIEWs: replace one definition, the entire chain updates. Downside: full query now takes longer since nothing is materialized.
:::

---

## `VIEW` vs `CREATE TABLE ... AS`

- `VIEW` = lazy (re-executed on every access)
- `CREATE TABLE ... AS` = **materialized** (stored on disk):

```{.sql}
CREATE TABLE wald_kantone_mat AS           -- <1>
SELECT
  name,
  st_area(st_intersection(w.geom, k.geom)) AS wald_area
FROM wald2 w, kantone k
WHERE st_intersects(w.geom, k.geom);
```

1. This stores the result as an actual table on disk

- Trade-off:
  - **Materialized**: faster to query, takes disk space, does **not** auto-update
  - **VIEW**: no extra storage, always current, slower to query

## Import into R

- Connect, load spatial extension, read the `VIEW`:

```{r}
#| eval: false
library(duckdb)

con <- dbConnect(
  duckdb(),
  dbdir = "data/week4-exercises/wald-kantone.duckdb",
  read_only = TRUE
)

dbExecute(con, "LOAD spatial;")
kanton_frac <- dbReadTable(con, "kanton_frac")

dbDisconnect(con)
```

## Wrap-up

- DuckDB = **in-process OLAP** database with a spatial extension
- SQL lets you express spatial analyses declaratively
- **R-Tree indices** make spatial predicates fast
- **VIEWs** let you build analyses incrementally — easy to iterate and scale
- Results flow back into R for further analysis or visualization

