---
title: "Week 3: Vector Advanced" 
---

## Topological relations

- We briefly touched topological relations in @sec-spatiao-queries
- Topological relations describe the spatial relationships between objects. 
- They are also called *Binary topological relationships* or *binary predicates*
- They are logical statements (`TRUE`/`FALSE`) about the spatial relationships between two objects
- The two objects are defined by ordered sets of points (typically forming points, lines and polygons) in two or more dimensions [@egenhofer1990]. 


### Usage

- Topological relations can be used to subset or join spatial data
- For example:
  - Subsetting: *Return all rivers that flow through the canton of Zurich*
  - Joining: *For each of the following Cinemas, what is the nearest train station?*

### Named topological relations {#sec-named-topological-relations}

- The most common topological relations are offered as functions / tools in most GIS software
- In `sf`, the following topological functions are available:
  - `st_intersects()`
  - `st_disjoint()`
  - `st_touches()`
  - `st_crosses()`
  - `st_within()`
  - `st_contains()`
  - `st_contains_properly()`
  - `st_overlaps()`
  - `st_equals()`
  - `st_covers()`
  - `st_covered_by()`
  - `st_equals_exact()`
  - `st_is_within_distance()`
- These all work slightly differently, and are used in different contexts.
- For example: 
  - `st_covers()` returns `TRUE` if no points of `x` are outside `y`, and at least one point of `x` is inside `y`.
  - `st_touches()` returns `TRUE` if the geometries have at least one point in common, but their interiors do not intersect.
  

### Named topological relations

- Some of the relations mentioned above are symmetrical (the order of the geometries does not matter)
- For example, if `st_touches(x, y)` is `TRUE`, then `st_touches(y, x)` is also `TRUE`
- Others are not, meaning that the order of the geometries is important. 
- For example, `st_contains(x, y)` returns `TRUE` if `x` contains `y`, but `st_contains(y, x)` returns `FALSE`
- Some of the relations require additional arguments, such as `st_is_within_distance()`, which requires a distance argument
- You can find the full list of topological relations in the `sf` documentation using `?geos_binary_pred`


  
### Subset Examples

- Let's take the example of the playgrounds in Zurich and the public transport stops. 
- We can subset the playground data to only include playgrounds that are close (e.g. 100m) to public transport stops


```{r}
library(sf)
library(readr)
# source: https://www.stadt-zuerich.ch/geodaten/
playgrounds <- read_sf("data/week4-exercises/playgrounds.gpkg")
publictransport <- read_sf("data/week4-exercises/public_transport.gpkg")

# Using the shorthand notation
playgrounds_close <- playgrounds[publictransport,,op = st_is_within_distance, dist = 100]

```


```{r}
#| echo: false
#| label: fig-playground-publictransport
#| fig-cap: "Note that the playgrounds within 100m of public transport (red dots) are a *subset* of all the playgrounds"

library(tmap)  

tm_shape(publictransport) + tm_dots(shape = 2, col = "darkblue") +
tm_shape(playgrounds) + tm_dots(shape = 16, fill = "darkgreen", size = .5)  +
  tm_shape(playgrounds_close) + tm_dots(shape = 16, fill = "red", size = .5) +
  tm_add_legend(fill = c("darkgreen","red", "darkblue"), col = c("darkgreen", "red", "darkblue"), shape = c(16,16, 2),labels = c("Playgrounds", "Playgrounds within 100m", "Public Transport")) 


```



### Spatial join Example

- In a spatial join, we want to add information from one dataset to another based on their spatial relationship
- For example, we can add the name of the *nearest* public transport stop to each playground.
- For this, we will use the function `st_nearest_feature`
- Strictly speaking, this function is not *binary predicate*, but is very useful for spatial joins
- To make the example clearer, I will first discard all unnecessary columns from the datasets

```{r}

publictransport <- publictransport[,"CHSTNAME"]

playgrounds <- playgrounds[, "name"]


playgrounds_join <- st_join(
  playgrounds, 
  publictransport, 
  join = st_nearest_feature
  )


playgrounds_join

```


### Spatial join Example (2)

- In the previous example, `playgrounds_join` has the same features as `playgrounds`, but with an additional column (`CHSTNAME`)
- The reason for this, is that there is only one *nearest station* for every playground
- In some joins, the number of rows in the joined dataset can be different from the original dataset
- For example, if we want to join via the method *within 100m*, the result is different (see below). 
- This is because there can be multiple public transport stops within 100m of a playground


```{r}
#| collapse: true
playgrounds_join2 <- st_join(
  playgrounds, 
  publictransport, 
  join = st_is_within_distance, 
  dist = 100
  )

nrow(playgrounds)

nrow(playgrounds_join2)

playgrounds_join2
```


### Spatial join order

- Note that as is the case with all joins, the order of the datasets matters
- `st_join` by default is a left join (the first dataset is the left dataset)
- The resulting dataset will have the geometry of the left dataset
- This is especially noticeable when joining datasets of different types

```{r}

kreise <- read_sf("data/week4-exercises/stadtkreise-zh.gpkg")

kreise <- kreise[,"STADTKREIS"]


publictransport_join <- st_join(publictransport, kreise)
```

```{r}
#| echo: false
tm_shape(publictransport_join) + tm_dots(fill = "STADTKREIS")
```

- if you reverse the order, each `stadtkreis` will be duplicated for every point in the `publictransport` dataset it intersects

```{r}
#| collapse: true

kreise_join <- st_join(kreise, publictransport)

nrow(kreise)

nrow(kreise_join)
```


```{r}
#| echo: false
#| label: fig-publictransport-kreise
#| fig-cap: "Note that all the duplicate stadtkreise overlap each other, so when you visualize the data, the issue is not noticeable"
#| fig-subcap: 
#|   - "Dataset `kreise` with 12 features"
#|   - "Dataset `kreise_join` with 477 features"
#| layout-nrow: 1
tm_shape(kreise) + tm_polygons("STADTKREIS") + tm_layout(legend.show = FALSE)

tm_shape(kreise) + tm_polygons("STADTKREIS") + tm_layout(legend.show = FALSE)

```


### Advanced topological operations

- In case the named topological relations (see @sec-named-topological-relations) do not covery a specific usecase, we can gain more fine grained control using so called *DE-9IM strings*. 
- DE-9IM stands for *Dimensionally Extended nine-Intersection Model* and is the formal definition of the topological relations between two geometries
- DE-9IM powers all binary predicate functions like `st_intersects` etc.
- The concept is a bit complex, but the idea is that it describes the topological relationship between two geometries in a 3x3 matrix. 
- The reasons the matrix is 3x3 is because it considers the intersection of the (1) interior, (2) boundary, and (3) exterior of the geometries.
- @tbl-de9im shows how two overlapping polygons are analyzed using DE-9IM 



```{r}
#| echo: false
#| label: tbl-de9im
#| tbl-cap: "Image source: M. W. Taves [commons.wikimedia.org](https://commons.wikimedia.org/wiki/File:DE9IM_II.svg)"


library(dplyr)
library(tidyr)


perspec <- c("Interior", "Boundary", "Exterior")
perspec2 <- c("I","B","E")

df <-cbind(rep(perspec2, each = 3), rep(perspec2, 3)) |> 
  apply(1, \(x)paste(x,collapse = "")) |> 
  (\(x)file.path("images","DE9IM",paste0("DE9IM_",x,".svg.png")))() |> 
  matrix(ncol = 3, byrow = TRUE) |> 
  as.data.frame()

colnames(df) <- perspec
rownames(df) <- perspec

library(gt)
df |> 
  gt::gt(rownames_to_stub = TRUE) |> 
  fmt_image(columns = c(Interior,Boundary,Exterior),height = "4em") 
  
```

  
- In the example in @tbl-de9im
  - the intersection of both *interiors* results in an area This is denoted as `2`
  - the intersection of *interior* with *boundary* results in a line, this is denoted as `1`
  - the intersection of *boundary* with *boundary* results in two points, this is denoted as `0`
- Once all nine intersections have been regarded and encoded, the result is flattened into a string (with nine characters) which describes the topological relationship precisely
- In the example in @tbl-de9im, a compact representation as string code is `212101212`
- This concept is especially interesting from the opposite perspective: i.e. we describe the DE-9IM string we are looking for (see next chapter)


## DE-9IM spatial predicates


https://r.geocompx.org/spatial-operations#DE-9IM-strings

```{r}
xy2sfc = function(x, y) st_sfc(st_polygon(list(cbind(x, y))))
x = xy2sfc(x = c(0, 0, 1, 1, 0), y = c(0, 1, 1, 0, 0))

grid = st_make_grid(x, n = 3)
plot(grid)
```






## Spatial data operations

see https://r.geocompx.org/spatial-operations#introduction-1




## DuckDB

DuckDB is an in-process SQL OLAP database management system. It is designed to be used as an embedded database in applications, but it can also be used as a standalone SQL database.

- In-process SQL means that DuckDB’s features run in your application, not an external process to which your application connects. In other words: there is no client sending instructions nor a server to read and process them. SQLite works the same way, while PostgreSQL, MySQL…, do not.

- OLAP stands for OnLine Analytical Processing, and Microsoft defines it as a technology that organizes large business databases and supports complex analysis. It can be used to perform complex analytical queries without negatively affecting transactional systems.

DuckDB is a great option if you’re looking for a serverless data analytics database management system.
